Early environment and numerical discrimination in a lizard

### 2024-09-22 ms.qmd
Updates

---
title: "The effects of prenatal corticosterone and thermal environment on numerical discrimination abilities in a lizard"
docx: 
  reference-doc: "./bib/tmpl.docx"
bibliography: "./bib/refs.bib"
csl: "./bib/behavioral-ecology-and-sociobiology.csl"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

Pablo Recio^1^,$\ddagger$ , Dalton C. Leibold^1^, Ondi L. Crino ^1,2^, Chris Friesen^3^, Amelia Y. Peardon ^1^, Daniel W.A. Noble^1^

@@

Pablo Recio ORCID: 0000-0002-5890-0218  
Dalton C. Leibold ORCID: 0000-0001-9645-2033  
Ondi L. Crino ORCID: 0000-0001-5700-1387  
Chris Friesen ORCID: 0000-0001-5338-7454  
Amelia Y. Peardon ORCID: 0009-0005-6227-8379  
Daniel W.A. Noble ORCID: 0000-0001-9460-8743

@@

## Abstract
Animals use quantitative judgements in several ecological contexts. From social behaviour to navigating their own environment, quantitative abilities can be crucial in animals' behaviour. Since prenatal conditions are predicted to impact brain development, quantitative competence may be as well affected by the early environment. Factors like temperature or stress-related hormones transmitted by the parents can play a prominent role in shaping cognitive traits. In adition, abrupt temperature changes can lead to the transmittion of stress-related hormones to the offspring, making essential to study the combined effects of stress hormones and thermal conditions during development. Here, we investigated the effects of prenatal corticosterone (CORT) and temperature on numerical discrimination in the Common garden skink (_Lampropholis guichenoti_). We manipulated CORT levels and incubation temperature of eggs employing a 2x2 factorial design, and then assessed numerousness through a spontaneous choice test with food as the relevant stimulus. Animals were subjected to five different tests with differen number of items and ratios (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 2 VS 4) in a repeated measures design. We hypothesized that CORT exposure and cold temperatures would constraint performance in the numerical task. Contrary to our expectations, we did not find any evidence of numerical discrimination for any of the treatments. These findings suggest that _L. guichenoti_ may not use numerical cues to make foraging decisions, and that early conditions does not affect their behaviour towards different quantities of food.

## Significant statement
Discriminating between quantities is crucial for animal behaviour, yet the impact of early environmental factors on the development of quantitative competence remains largely unexplored. In this study, we investigated for the first time the effects of prenatal corticosterone exposure and incubation temperature on numerical discrimination in a reptile. Contrary to our predictions, we did not observe any evidence of numerical discrimination regardless of incubation conditions. Our results align with previous studies showing that reptiles consuming live prey struggle to differentiate prey numbers in spontaneous choice tests. These results open new avenues for exploring the evolution of numerical abilities in reptiles and identifying factors influencing the development of quantitative skills in this group. Further research should explore alternative methods, such as trained tasks, to better understand the cognitive mechanisms underlying numerical competence in reptiles.

## Introduction  
The ability to discern between more and fewer quantities is a basic cognitive skill widely observed across the animal kingdom [@stancher2015numerical; @vonk2012bears; @beran2016capuchin]. It is considered adaptive in various ecological contexts, contributing to behaviors such as social decisions, mate selection, or foraging [@lucon2017individual; @nieder2018evolution]. For instance, female lions (_Panthera leo_) use numerical information to assess the risk of confronting rival groups [@mccomb1994roaring], while other animals use these abilities for navigating their environment [@chittka1995can], locate potential mates [@carazo2009quantity], or to make optimal foraging decisions [@uller2003salamanders]. Given the adaptive advantages of numerical and quantity discrimination, research has extensively explored the ability of different species to differentiate between quantities under different contexts and paradigmas.

Numerical competences are often evaluated by considering the types of cues animals use [@plotnik2019elephants; @cooper2024tell], the ecological contexts in which these abilities are deployed [@mccomb1994roaring; @chittka1995can; @uller2003salamanders; @carazo2009quantity], and the neural mechanisms that underlie them [@nieder2005counting]. The later have demonstrated that non-verbal numerical representations in mammals activate an extensive cortical network, with the prefrontal cortex and the intraparietal sulcus being particularly relevant [@nieder2005counting]. This connection between numerosity and brain physiology underscores the importance of understanding how brain development influences numerical abilities.

Animal's brain is particularly sensitive to environmental inputs during early stages of development [@coomber_independent_1997; @zhu_prenatal_2004], with long-lasting effects on cognitive abilities [@zhu_prenatal_2004; @amiel_egg_2014]. For instance, increases in glucocorticoids (GCs) - the main groups of hormones mediating the response to stressors [@sapolsky_how_2000] - during early life can alter animals' performance in spatial, associative, or reversal learning tasks [@szuran_water_1994; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @lui2017chronic]. These effects have been shown to vary depending on factors such as dose, subjects' sex, or type of cognitive test employed [@szuran_water_1994 @farrell_developmental_2015-learn; @bebus_associative_2016; @haq_prenatal_2021], but little is known about how exposure to GCs during development can affect quantitative abilities. At the same time, thermal early environment plays a prominent role in brain development and cognitive abilities in ectotherms [@coomber_independent_1997;@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_incubation_2018; @vila_pouca_quantity_2019]. For example, Port Jackson shark incubated at elevated-temperatures learned faster in a quantity discrimination task [@vila_pouca_quantity_2019]. However, the effects of early thermal environment on numerical abilities in other taxa remains largely unexplored.

Given that abrupt temperature changes can increase GC levels in vertebrates, leading to elevated GCs in offspring [@racic_effects_2020; @Crino_2023], it is essential to study the combined effects of early-life GCs and thermal conditions on cognitive development. Understanding how these factors influence quantitative abilities could reveal important insights into the ecological relevance of numerosity. To fill this gap, we used the common garden skink (_Lampropholis guichenoti_) as a model to test the effects of prenatal corticosterone (CORT) - the main GC in reptiles [@Crino_2023] - exposure and thermal environment on numerical discrimination abilities.

Traditionally, research on numerical abilities has been mainly focused on mammals, birds, and fishes [@nieder2005counting; @agrillo2014spontaneous], but recent studies demonstrate that reptiles are also competent in numbers. For example, red-footed turtles (_Chelonoidis carbonaria_) rapidly learned to associate colour to higher volumes of food [@soldati2017long], Hermann’s tortoises (_Testudo hermanni_) proved to discriminate between big and small food quantities in a spontaneous choice test [@gazzola2018continuous], and freshwater turtles (_Trachemys scripta_ and _Mauremys sinensis_) showed high skills when tested in a trained quantity discrimination task [@lin2021superior; @lin2024trained]. In lizards, __Podarcis siculus_ failed to discriminate between different number of food items in a spontaneous choice test [@miletto2017quantitative], but succeeded in a trained quantity discrimination task [@miletto2018quantity], whereas rock lizards (_Iberolacerta cyreni_) exhibited a clear preference for larger quantities of crickets when presented 5 versus 10 items [@recio_prey_2021]. Inded, the literature evidences basic quantitative abilities in reptiles, that may vary with the type of cue and test employed.

At the same time, reptiles' cognition is known to be significantly influenced by the environment in which they develop. Incubating lizards at elevated temperatures has been shown to improve cognitive abilities in consistency with elevated neuronal density in some areas of the brain [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017]; and thermal environment impaired cognition only under incubation temperatures beyond lizards' natural range [@dayananda_incubation_2017 or @abayarathna_effects_2020]. The effects of prenatal CORT on cognitive abilities in reptiles is less understood, but in other taxa, high levels of GCs are tipically related to cognitive impairments [@szuran_water_1994; @zhu_prenatal_2004].  However, the combined influence of prenatal CORT and thermal environment remains unknown, particularly on numerical abilities.

Our study aimed to address this gap by manipulating CORT concentration in _L. guichenoti_ eggs and incubating them under one of two temperature regimes (Cold -- 23ºC ± 3ºC or Hot -- 28ºC ± 3ºC) in a 2x2 factorial design. After hatching, we assessed the lizards' numerical discrimination abilities in a spontaneous choice test, where they were presented with different numbers of crickets. Based in other studies that explored the cognitive effect of prenatal GCs or thermal early environment in reptiles and other taxa, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. Additionally, we predicted that increased CORT levels would impair numerical competence [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures temperatures. Consistent with Weber's law [@agrillo2014spontaneous], we expected that lizards incubated at higher temperatures would successfully discriminate between quantities even at high ratios (e.g. 3 VS 4), whereas event the lowest ratios (e.g. 1 VS 4) could be challenging for those exposed to CORT and/or incubated at lower temperatures.

@@

#### Husbandry  
*Breeding colony* -- Lizards tested in out experiment came from a breeding colony established in the lab since 2019. The colony consists of approx. 180 adults of and _L. guichenoti_ housed in commonal containers (41.5 L x 30.5 W x 21 H cm) with six lizards (2 males and 4 females) per enclosure. Enclosures are provided with non-stick matting, shelter, and several small water dishes filled daily; and lizards are fed mid-size crickets (*Acheta domestica*) three days a week. Crickets are dusted with calcium weekly and multivitamin and calcium biweekly. Room temperatures are set to 22-24ºC, but to ensure a temperature gradient, we employ a heat chord and a heat lamp following a 12 h light:12 h dark cycle, keeping the warm side of enclosures at 32ºC.

*Eggs collection and incubation* -- Between mid-October 2023 to the end of February 2024, we placed small boxes (12.5 L x 8.3 W x 5 H cm) with moist vermiculite inside in one extreme of the communal enclosures (see above) to provided females with a place to lay the eggs. We checked for the presence of eggs in the boxes three days a week. Eggs' length and width were measured with a digital caliper to the nearest 0.1 mm, while mass was recorded with a (OHAUS, Model spx123) digital scale ± 0.001g error. Eggs were assigned a clutch number and a individual identity. Then we treated eggs with CORT or vehicle (see CORT and Temperature manipulation below) and placed them in individual cups (80 mL) with moist vermiculite (12 parts water to 4 parts vermiculite). We covered the cups with cling wrap to retain moisture and then left the eggs in LATWIT 2X5D-R1160 incubators at two different temperatures (see CORT and Temperature manipulation below) until hatching.

*Hatchlings* -- Incubators were checked three times a week for the presence of hatchlings. Once hatchlings were collected, we measured SVL and Tail Length (TL) with a ruler to the nearest mm, and weighed them with a (OHAUS, Model spx123) digital scale ± 0.001g error. Hatchlings were then placed in individual enclosures (18.7L x 13.2W x 6.3H cm) provided with non-stick matting and a small water dish. During this period, they were sprayed water every day and received 3-6 small *A. domestica* crickets three times a week. All care otherwise follows similar protocols to adults (see above).

@@

We manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (Cold -- 23ºC ± 3ºC or Hot -- 30ºC ± 3ºC) in a 2x2 factorial design (@fig-Methods A). We first allocated eggs onto one of two different treatments: in the CORT treatment, eggs were topically supplied with 5µL of CORT dissolved in 100% Ethanol (vehicle) at a final (10 pg CORT/mL) concentration; in the Control treatment eggs received an equal volume of the vehicle. CORT concentration employed in the CORT treatment represents 2 standard deviations above the mean natural concentration obtained in eggs from both species (non-published data). Then, eggs were incubated in one of the two previously mentioned temperature regimes (Cold or Hot) until hatching. This temperatures represent the upper and lower limit of the natural incubation temperatures [@qualls2000post]. The number of eggs per clutch assigned to each hormone and temperature treatment were counterbalanced in both species.

@@

Two weeks before starting the experimen phase (see below), lizards were moved to the experimental arena for acclimatation. The arenas were individual medium size (41 L x 29.7 W x 22 H cm) plastic containers provided with a shelter (9 L x 6 W x 1.5 H cm) on one of the extremes and a water dish on the center (@fig-Methods B). The new enclosures were placed in 7 different racks located in two different rooms. Each rack was associated to its own CCTV systems (device model DVR-HP210475) allowing us to record lizards' behaviour during the experiment. The number of lizards per species and treatment in each rack was counterbalanced to control for any effect of the room or the position of the lizard in the rack. During acclimatation and throughout the experiment, lizards were fed with only one cricket per day dusted with calcium and multivitamin (see protocol below), and water was supplied *ad libitum*. The temperature of both rooms was set to between 22-24ºC, but we also provided a temperature gradient by means of a heat cord along the center of the arena, and heat lamps in a 12 h light: 12 h dark cycle.

The numerical discrimination test consisted in a series of two-choice trials where lizards where presented with different number of food items. Each choice was placed in one of two sides of a white 3D printed platform (see @fig-Methods B, C) and then covered with a transparent Petri dish to prevent the lizard from obtaining the food. We performed five tests that differed in the ratio between both choices: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4. We never used more than four items because higher numbers are considered to be represented by a different system [@nieder2005counting; @cantlon2009beyond; @hyde2011two]. The food consisted in frozen small crickets (*A. domestica*) that were thawed and dusted with calcium and multivitamin. We decided to use frozen crickets to avoid movement to affect lizards' behaviour. Lizards were fed frozen crickets during one month before the beginning of the experiments for habituation to frozen prey. Two weeks before the experiments, to habituate lizards to the platform and the Petri dish, the frozen cricket that was placed on top of the Petri dish in one of the sides of the platform, simulating the conditions of the experiment (see @fig-Methods B). The side where the food was placed was changed every day to avoid any side bias during the trials. However, to further control for that, the side where we placed the higher number of crickets was randomised between trials and individuals. We also counterbalanced the amount of individuals per treatment tested in the same type of trial each day. The crickets used during the experiment were all the same size, but to control for other quantitative traits [@agrillo2014spontaneous], we changed the orientation of the crickets as shown in @fig-Methods D. In addition, after the experiments, we made another set of trials to test whether the orientation of the crickets could affect lizards' choice, but we found no effect of orientation (see Supplementary Material - Prey orientation test).

All tests were done between 10 and 12 in the morning when the lizards were most active. Each trial was performed daily for five days, between the 17^th^ to the 22^nd^ of May 2024. We recorded three main variables: i) Latency, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard, recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) Interest shown for the higher amount of food (detailed below). We considered that the lizard was interacting with the food when it touched the Petri dish with its snout or when the front half of the body was on top of the platform. For five minutes after the first interaction with one of the options, we recorded the time spent interacting with that option for a maximum of five minutes; we did this for both amounts of food independently of what was the first choice. We measured the interest shown for the higher amount of food by substracting the time interacting with the option with less crickets to the time spent interacting with the larger amount of food. If there was no interaction with one of the options during the test, we considered the time spent as 0 for that option. When there was no interaction with any of the choices, both Latency, First choice, and Interest were noted as NA. We also considered Interest as NA in some few cases where the animals managed to lift the dish and eat the prey; Latency and First choice were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests.

@@

#### Statistical analyses  
In our analyses, we fitted a multivariate Bayesian multilevel model using the brm function from the _brms_ package [@burkner2017brms] in R extension (version 2.8.2) [@R]. We ran four parallel MCMC chains of 4000 iterations with a warmup period of 1000 iterations. We modelled the three main recorded behaviour: Latency (log-transformed), Choice, and Interest (see above) as the response variables, and test (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4), hormone (CORT versus Control), incubation temperature (Cold versus Hot) and their interaction as fixed factors. We also included in our predictors sex, and age (centered to 0). We included lizard indentity and clutch as random factors. We used the posterior distributions of parameters from these models to test for differences in the ability to discriminate between different number of crickets between treatments and species, for each of the numerical tests. We calculated the contrast between treatments for each test using p~mcmc~ values, and we also estimated whether predicted Choice and Interest differed from 0.5 or zero respectively using p~mcmc~. We considered statistical significance if p~mcmc~ < 0.05.

#### Ethical note
Subjects were collected from a breeding colony kept in the lab since 2019. All animals used in this experiment were provided humane laboratory housing, with thermorregulation opportunity, light, and humidity conditions (see above for details). Lizards in the colony received five-seven crickets per individual three times a week while the subjects in the experiment were fed once per day. All lizards were handled minimally to avoid stress. Experimental procedures and laboratory housing complied with Australian law and were approved by the Australian National University Animal Experimentation Ethics Committee (A2022_33). Upon completion of experimental trials, lizards were kept in their enclosure and will be euthanized following standard humane procedures once they reach the human-ending point.

>>>>

---
title: "Early environment and numerical discrimination in a lizard"
docx: 
  reference-doc: "./bib/tmpl.docx"
bibliography: "./bib/refs.bib"
csl: "./bib/behavioral-ecology-and-sociobiology.csl"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

Pablo Recio^1^,$\ddagger$ , Dalton C. Leibold^1^, Ondi L. Crino ^1,2^, Christopher R. Friesen^3^, Amelia Y. Peardon ^1^, Daniel W.A. Noble^1^

@@

Pablo Recio ORCID: 0000-0002-5890-0218  
Dalton C. Leibold ORCID: 0000-0001-9645-2033  
Ondi L. Crino ORCID: 0000-0001-5700-1387  
Christopher R. Friesen ORCID: 0000-0001-5338-7454  
Amelia Y. Peardon ORCID: 0009-0005-6227-8379  
Daniel W.A. Noble ORCID: 0000-0001-9460-8743

@@

## Abstract
From social behaviour to navigating their environment, quantitative abilities can be crucial in animals' life. Early life conditions are predicted to impact brain development, and quantitative competence may not be exent of these effects. Early thermal environment or glucocorticoids (GCs) - stress-related hormones - transmitted by the parents can play a prominent role in shaping cognition through effects on brain physiology. Furthermore, temperature can lead to increased levels of GCs that are passed on to the offspring, emphasizing the need to study the combined effects of these two factors. Here, we investigated the effects of prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature on numerical discrimination in the common garden skink (_Lampropholis guichenoti_). We manipulated CORT levels and temperature during incubation and then assessed numerical abilities through a spontaneous choice test with food as the relevant stimulus. Employing a repeated measures design, we subjected lizards to five numerical tests each differing in the total number of items and the ratios between the two choices (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 2 VS 4). Contrary to our expectations, we found no evidence of numerical discrimination or any effect of treatments on lizards' behaviour. These findings suggest that _L. guichenoti_ may not use numerical cues to make foraging decisions and that early conditions do not affect their behaviour towards different quantities of food.

## Significance statement
Discriminating between quantities is crucial for animal behaviour, yet the impact of early environmental factors on the development of quantitative competence remains largely unexplored. In this study, we investigated for the first time the effects of prenatal corticosterone exposure and incubation temperature on numerical discrimination in a reptile. Contrary to our predictions, we did not observe any evidence of numerical discrimination regardless of incubation conditions. Our results align with previous studies showing that reptiles consuming live prey struggle to differentiate prey numbers in spontaneous choice tests. These results open new avenues for exploring the evolution of numerical abilities in reptiles and identifying factors influencing the development of quantitative skills in this group. Further research should explore alternative methods, such as trained tasks, to better understand the cognitive mechanisms underlying quantitative competence in reptiles.

## Introduction  
The ability to discern between quantities is a fundamental cognitive skill widely observed across the animal kingdom [@stancher2015numerical; @vonk2012bears; @beran2016capuchin]. Numerical cognition is adaptive in various ecological contexts, contributing to social decisions, mate selection, or foraging [@lucon2017individual; @nieder2018evolution]. For instance, female lions (_Panthera leo_) use numerical information to assess the risk of confronting rival groups [@mccomb1994roaring], male adult mealworms (_Tenebrio molitor_) select sites with scents of more females [@carazo2009quantity], and red-backed salamanders (_Plethodon cinereus_) prefer tubes containing larger numbers of prey [@uller2003salamanders]. Given the adaptive advantages of numerical and quantity discrimination, research has extensively explored the ability of different species to differentiate between quantities under different contexts and paradigms.

Numerical competence is often evaluated by considering the types of cues animals use [@plotnik2019elephants; @cooper2024tell], the ecological contexts in which these abilities are deployed [@mccomb1994roaring; @chittka1995can; @uller2003salamanders; @carazo2009quantity], and the neural mechanisms that underlie them [@nieder2005counting]. The latter have demonstrated that non-verbal numerical representations in mammals activate an extensive cortical network, with the prefrontal cortex and the intraparietal sulcus being particularly relevant [@nieder2005counting]. This connection between numerosity and brain physiology underscores the importance of understanding how brain development influences numerical abilities.

Animals' brains are particularly sensitive to environmental inputs during early stages of development [@coomber_independent_1997; @zhu_prenatal_2004], with long-lasting effects on cognitive abilities [@zhu_prenatal_2004; @amiel_egg_2014]. For instance, glucocorticoids (GCs), the main groups of hormones mediating the physiological stress response [@sapolsky_how_2000], can alter animals' performance in spatial, associative, or reversal learning tasks if elevated during early stages of development [@szuran_water_1994; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @lui2017chronic]. Prenatal exposure to GCs is typically associated with impaired learning skills, but these effects have been shown to vary depending on factors such as dose, subjects' sex, or type of cognitive test employed [@szuran_water_1994; @farrell_developmental_2015-learn; @bebus_associative_2016; @haq_prenatal_2021]. Nevertheless, little is known about how exposure to GCs during development can affect quantitative abilities.

At the same time, the early thermal environment plays a prominent role in brain development and cognitive abilities in ectotherms [@coomber_independent_1997;@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_incubation_2018; @vila_pouca_quantity_2019]. For example, Port-Jackson sharks (_Heterodontus portusjacksoni_) incubated at elevated temperatures learned faster in a quantity discrimination task [@vila_pouca_quantity_2019]. However, the effects of early thermal environment on numerical abilities in other taxa remains largely unexplored.

Given that abrupt temperature changes can increase GC levels in vertebrates, leading to elevated GCs transmitted to the offspring [@racic_effects_2020; @Crino_2023], it is essential to study the combined effects of early-life GCs and thermal conditions on cognitive development. Understanding how these factors influence quantitative abilities could reveal important insights into the ecological relevance of numerosity. To fill this gap, we used the common garden skink (_Lampropholis guichenoti_) as a model to test the effects of elevated prenatal corticosterone (CORT) - the main GC in reptiles [@Crino_2023] - and incubation temperature on numerical discrimination abilities.

Traditionally, research on numerical abilities has been mainly focused on mammals, birds, and fishes [@nieder2005counting; @agrillo2014spontaneous], but recent studies demonstrate that reptiles are also competent in numberical abilities. For example, red-footed turtles (_Chelonoidis carbonaria_) rapidly learned to associate colour with higher volumes of food [@soldati2017long], Hermann’s tortoises (_Testudo hermanni_) proved to discriminate between large and small food quantities in a spontaneous choice test [@gazzola2018continuous], and freshwater turtles (_Trachemys scripta_ and _Mauremys sinensis_) rapidly learned a quantity discrimination task [@lin2021superior; @lin2024trained]. The italian wall lizard (_Podarcis siculus_) failed to discriminate between different numbers of food items in a spontaneous choice test [@miletto2017quantitative], but succeeded in a trained quantity discrimination task [@miletto2018quantity]. In contrast, Iberian rock lizards (_Iberolacerta cyreni_) preferred larger quantities of crickets when presented five versus ten items [@recio_prey_2021]. Indeed, the literature shows evidence for basic quantitative abilities in reptiles that may vary with the type of cue and the test employed.

Reptiles' cognition is significantly influenced by the environment in which they develop. For example, incubating lizards at elevated temperatures improves cognitive abilities consistent with elevated neuronal density in some areas of the brain [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017], but impairs cognition at temperatures above lizards’ natural incubation thermal range [@dayananda_incubation_2017; @abayarathna_effects_2020]. The effects of prenatal CORT on cognitive abilities in reptiles is less understood, but in other taxa, high levels of GCs are typically related to cognitive impairments [@szuran_water_1994; @zhu_prenatal_2004]. However, the combined influence of prenatal CORT and thermal environment on cognition remains unknown, particularly on numerical abilities.

Our study aimed to address this question by manipulating CORT concentration (CORT-treated or a sham control) in _L. guichenoti_ eggs and incubating them under one of two temperature regimes (Cold - 23 ± 3ºC or Hot - 28 ± 3ºC) in a 2x2 factorial design. After hatching, we assessed the lizards' numerical discrimination abilities in a spontaneous choice test, where they were presented with different numbers of crickets. Based on other studies that explored the cognitive effects of prenatal GCs or thermal early environment in reptiles and other taxa, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. Additionally, we predicted that increased CORT levels would impair numerical competence [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures. In this sense, a common observation in nature is that the ability to detect a change in stimulus intensity depends on the relative, not absolute, difference between the stimuli ['Weber's law'; @agrillo2014spontaneous]. Consistent with Weber's law, we predicted that lizards incubated at higher temperatures would successfully discriminate between quantities even at high ratios (e.g. 3 VS 4), whereas even the lowest ratios (e.g. 1 VS 4) could be challenging for those exposed to CORT or incubated at lower temperatures.

@@

#### Animal husbandry  
*Breeding colony* -- The lizards tested in our experiment came from a breeding colony established in the lab in 2019. The colony consisted of approximately 180 adults of _L. guichenoti_ housed in communal containers (41.5 L x 30.5 W x 21 H cm) with six lizards (2 males and 4 females) per enclosure. Enclosures were provided with non-stick matting, shelter, and several small water dishes filled daily, and lizards were fed mid-size crickets (*Acheta domestica*) three days a week. Crickets were dusted with calcium weekly and multivitamins and calcium biweekly. Room temperatures were set to 22-24ºC, but to ensure a temperature gradient, we employed a heat chord and a heat lamp following a 12 h light:12 h dark cycle, keeping the warm side of enclosures at 32-34ºC.

*Egg collection and incubation* -- Between mid-October 2023 to the end of February 2024, we placed small boxes (12.5 L x 8.3 W x 5 H cm) containing vermiculite on one side of the communal enclosures (see above) to provide females with a place to lay the eggs. We checked for eggs in the boxes three days a week. Eggs' length and width were measured with a digital calliper to the nearest mm (± 0.1 mm), while mass was recorded with a digital scale (OHAUS, Model spx123) to the nearest g (± 0.001 g). Eggs were assigned a clutch number and an individual identity. Then we treated eggs with CORT or vehicle (see CORT and Temperature manipulation below) and placed them in individual cups (80 mL) with moist vermiculite (12 g water to 4 g vermiculite). We covered the cups with plastic wrap to retain moisture and left the eggs in incubators (LATWIT 2X5D-R1160) programmed to two different thermal regimes (see CORT and Temperature manipulation below) until hatching. Incubators were checked for the presence of hatchlings three times a week.

*Hatchlings* -- Immediately after hatching, we measured Snout-Vent Length (SVL) and Tail Length (TL) with a ruler to the nearest mm (± 0.1 mm), and mass using a digital scale (OHAUS, Model spx123) to the nearest g (± 0.001 g). Hatchlings were then placed in individual enclosures (18.7L x 13.2W x 6.3H cm) provided with non-stick matting and a small water dish. During this period, they were given water daily and received 3-6 small *A. domestica* crickets three times a week. All care otherwise follows similar protocols to adults (see above).

@@

We manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (Cold -- 23 ± 3ºC or Hot -- 30 ± 3ºC) in a 2x2 factorial design (@fig-Methods A). Eggs were allocated to one of two different developmental stress treatments: CORT-treated eggs were topically supplied with 5µL of CORT dissolved in 100% Ethanol at a final concentration of 10 pg CORT/mL. In the Control treatment eggs received an equal volume of the 100% Ethanol vehicle. We selected those doses based on previous studies where CORT treatment increased mean yolk CORT levels ~3.7x higher than control eggs in _Lampropholis delicata_ [@crino2024eggs]. Then, eggs were incubated in one of the two previously mentioned temperature regimes (Cold or Hot) until hatching. These temperatures represent the upper and lower limit of the natural incubation temperatures for this species [@qualls2000post]. The number of eggs per clutch assigned to each hormone and temperature treatment was counterbalanced per clutch.

@@

Two weeks before starting the numerical discrimination task (see below), lizards were moved to the experimental arenas for acclimatisation. The arenas were individual medium-size (41 L x 29.7 W x 22 H cm) plastic containers provided with a shelter (9 L x 6 W x 1.5 H cm) on one of the sides and a water dish in the center (@fig-Methods B). These arenas were placed in 7 different racks located in two experimental rooms. Each rack was associated with its own CCTV system (device model DVR-HP210475), allowing us to record lizards' behaviour during the experiment. Despite the conditions in the experimental rooms were the same as in the colony room, we counterbalanced the number of lizards per treatment in each rack to control for any potential effects of the room or the position of the lizard in the rack. During acclimatisation and throughout the experiment, lizards were fed with only one cricket daily, dusted with calcium and multivitamins (see protocol below), and water was supplied *ad libitum*. The temperature of both rooms was set to 22-24ºC, but we also provided a temperature gradient (24-32°C) using a heat cord along the center of the arena and heat lamps in a 12 h light: 12 h dark cycle.

The numerical discrimination test consisted of a series of two-choice trials where lizards were presented with different numbers of food items. Each choice was placed on one of the two lateral platforms of a white 3D printed device (see @fig-Methods B, C) and then covered with a transparent Petri dish to prevent the lizard from obtaining the food. We performed five tests that differed in the ratio between both choices: 1 cricket VS 4 crickets, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4. The food consisted of small frozen crickets (*A. domestica*) that were thawed and dusted with calcium and multivitamins. We decided to use frozen crickets to avoid movement that would affect lizards' behaviour [@cooper2024tell].

Lizards were fed frozen crickets for one month prior to the experiments to habituate them to eating frozen prey. To habituate lizards to the experimental conditions, frozen crickets were placed on top of one of the platforms and Petri dishes during each feeding for two weeks prior to experimental trials. The side where the food was placed was changed every day to avoid any side bias during the trials. However, to further control for that, the side where we placed the higher number of crickets was randomised between trials and individuals during the tests. We also counterbalanced the number of individuals per treatment tested in the same type of trial each day. The crickets used during the experiment were all the same size, but to control for other quantitative traits [@agrillo2014spontaneous], we changed the orientation of the crickets as shown in @fig-Methods D. In addition, after the experiments, we conducted another set of trials to test whether the orientation of the crickets could affect lizards’ choice, but we found no effect of orientation (see Supplementary Material - Prey orientation test).

@@

All tests were done between 1000-1200 hours when the lizards were most active. Each trial was performed daily for five days between the 17^th^ and 22^nd^ of May 2024. We recorded three main variables: i) Latency, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard (Choice), recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) Interest shown for the higher amount of food (detailed below). We considered that the lizard interacted with the food when it touched the Petri dish with its snout or when the front half of the body was on top of the platform. For five minutes after the first interaction with one of the options, we recorded the time spent interacting with that option for a maximum of five minutes; we did this for both amounts of food independently of what was the first choice. We measured the interest shown for the higher amount of food (Interest) by substracting the time interacting with the option with fewer crickets to the time spent interacting with the larger amount of food. If there was no interaction with one of the options during the test, we considered the time spent as 0 for that option. When there was no interaction with any of the choices, both Latency, Choice, and Interest were noted as NA. We also considered Interest as NA in a few cases where the animals managed to lift the dish and eat the prey; Latency and Choice were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests.

#### Statistical analyses  
In our analyses, we fitted a multivariate Bayesian multilevel model using the brm function from the _brms_ package [@burkner2017brms] in R extension (version 2.8.2) [@R]. We ran four parallel MCMC chains of 3000 iterations for each model, with a warmup period of 1000 iterations. We checked that all MCMC chains converged (Rhat < 1.2) and were mixing effectively to ensure we had >1000 effective samples from the posterior distribution.

We modelled the three main recorded behaviours: Latency, Choice, and Interest as the response variables. We assumed normal distributions for Latency (log-transformed) and Interest [family = gaussian()]. The error structure for Choice was modeled using a Bernoulli distribution with a logit link function [family = Bernoulli(link = ‘logit’)].  We included test (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4), hormone (CORT versus Control), incubation temperature (Cold versus Hot), and the three-way interaction as fixed factors. We also included in our predictors sex, and age (centered to 0). We included lizard identity and clutch as random factors. _L. guichenoti_ lays up to two clutches per year [@joss1985reproductive]. Since egg collection was done during half of the breeding season, each clutch likely came from a unique mother, and so, clutch identity captures potential maternal effects.

We used the posterior distributions of parameters from these models to test for between treatment differences in the ability to discriminate different numbers of crickets for each of the numerical tests. We used p~mcmc~ to tests whether the contrast between treatments for each test differed from zero and whether Choice and Interest differed from 0.5 or zero respectively. We considered statistical significance if p~mcmc~ < 0.05.

#### Ethical note
Subjects were collected from a breeding colony that had been kept in the lab since 2019. All animals in this experiment were provided humane laboratory housing, thermoregulation opportunities, and light and humidity conditions (see above for details). Lizards in the colony received five to seven crickets per individual three times a week while the subjects in the experiment were fed once per day. All lizards were handled minimally to avoid stress. Experimental procedures and laboratory housing complied with Australian law and were approved by the Australian National University Animal Experimentation Ethics Committee (A2022_33). Upon completion of experimental trials, lizards were kept in their enclosure and will be euthanized following standard humane procedures once they reach the humane endpoint.
